Figures

### 2016-01-03 scripts/figures.Rmd
ordinations

---
title: "Figures"
author: "JT McCrone"
date: "December 30, 2015"
output: html_document
---
```{r,settings_and_lib}
require(ggplot2)
require("plyr")
require("Biostrings")
require(reshape2)
require("pROC")
require(gridExtra)
theme_set(new = theme_classic(base_size=24))
```

## 1) Miseq out of the box
### ROC skematic
```{r, roc_skematic}
set.seed(42) # make the figures reproducible
example<-data.frame(Condition=c(rep(x = c(TRUE,FALSE),each=20)),test=c(runif(20,-10,-4),runif(20,-6,0))) # Make a dataframe of T and F positives with random p values that favors a lower distribution for the TP
ex.roc<-roc(example$Condition~example$test) # make an ROC based on this data

@@

# Plot the roc to be used later
roc.p<-ggplot(roc.df,aes(x=1-specificity,y=sensitivity))+geom_step(color="black",size=2)+xlab("1-Specificity")+ylab("Sensitivity")+theme(axis.title.y=element_text(vjust=1))

# png("../../Talks/figures/roc.ex.png")
# roc.p
# dev.off()

# plot the point distribution to be used later
points.p<-ggplot(example,aes(x=1,y=test,fill=Condition))+geom_dotplot(binaxis = "y", stackdir = "center")+scale_fill_manual(values=c("white","black"),name="",breaks=c(T,F),labels=c("True Positive","False Positive"))+xlab("")+ylab("Log(p value)")+theme(axis.ticks = element_blank(), axis.text.x = element_blank())

#+scale_fill_brewer(palette="Set1")#+theme(text = element_text(size=30))

#This function takes in a specificity and adds lines indicating where this specificity is found on the ROC plotted above
plot.roc<-function(x){ 
  above.sense.cut<-subset(roc.df,specificity>=x,select=c(threshold,specificity,sensitivity)) # get the points where specificity is above the cutoff
  sense.min<-min(above.sense.cut$specificity) # get the minimum of these to get as close to desired sensitivity as possible
  sense.cut<-subset(above.sense.cut,specificity==sense.min) # get all rows where sensitivity is at this point
  spec.max<-max(sense.cut$sensitivity) # but there can be different FPR for each TPR so here we maximize specificity
  output<-subset(sense.cut,sensitivity==spec.max)
  lines<-data.frame(sense=c(output$sensitivity,output$sensitivity,0), # make a data frame with the coordinates for the lines whicha re to be added
                    spec=c(0,1-output$specificity,1-output$specificity))
  p1<-roc.p+geom_line(data=lines,aes(x=spec,y=sense),size=1.5,lty=2)+xlab("1-Specificity")+ylab("Sensitivity") # plot

  p1<-p1+theme(axis.title.y = element_text(vjust=1.2))+theme(axis.title.x = element_text(vjust=0))
  print(p1)
}

#This function takes in a specificity and adds a line to show where this specificity is reached in the p value distribution above

plot.pv<-function(x){
  above.sense.cut<-subset(roc.df,specificity>=x,select=c(threshold,specificity,sensitivity)) # get the points where sensitivity is above the cutoff
  sense.min<-min(above.sense.cut$specificity) # get the minimum of these to get as close to desired sensitivity as possible
  sense.cut<-subset(above.sense.cut,specificity==sense.min) # get all rows where sensitivity is at this point
  spec.max<-max(sense.cut$sensitivity) # but there can be different FPR for each TPR so here we maximize specificity
  output<-subset(sense.cut,sensitivity==spec.max)
  p2<-points.p+geom_hline(yintercept=output$threshold,
                          size=1.5,lty=2)
  p2<-p2+scale_x_continuous(breaks=c(1),labels=c(""))
  p2<-p2+xlab("")+ylab("Log(p value)")
  p2<-p2+theme(axis.title.y = element_text(vjust=1.2))
  print(p2)
}

pdf("../results/figures/pv.1.pdf")
plot.pv(0.35)
dev.off()

pdf("../results/figuresROC.1.pdf")
plot.roc(0.35)
dev.off()

pdf("../results/figures/pv.2.pdf")
plot.pv(0.95)
dev.off()

pdf("../results/figures/ROC.2.pdf")
plot.roc(0.95)
dev.off()

```
Now it might also be useful to include a figure that has 2 cutoffs and outputs the plots side by side.

```{r,roc_skematic_onplot}
#Helper function that gets the threshold and sensitivity from and roc at a given speceificity
helper.roc<-function(spec,roc.df){
  above.sense.cut<-subset(roc.df,specificity>=spec,select=c(threshold,specificity,sensitivity)) # get the points where specificity is above the cutoff
  sense.min<-min(above.sense.cut$specificity) # get the minimum of these to get as close to desired sensitivity as possible
  sense.cut<-subset(above.sense.cut,specificity==sense.min) # get all rows where sensitivity is at this point
  spec.max<-max(sense.cut$sensitivity) # but there can be different FPR for each TPR so here we maximize specificity
  output<-subset(sense.cut,sensitivity==spec.max)
}

@@

### DeepSNV ROC

```{r,roc_helper_runctions}
id<-function(x){
  x<-strsplit(x,".",fixed=T)[[1]][1]
}

@@

adjust.coords<-function(roc.df,sum.df,possible_tp,possible_vars){ # adjust the sensitivity and specificity called by pROC  
  samp<-roc.df$samp[1]
  samp.df<-subset(sum.df,Id==samp)
  sense.factor<-length(which(samp.df$category==T))/possible_tp
  TN.samp<-length(which(samp.df$category==F))
  mutate(roc.df,adj.sensitivity=sensitivity*sense.factor,FP=(TN.samp-TN.samp*specificity),TP=sense.factor*possible_tp,adj.specificity=(possible_vars-possible_tp-FP)/((possible_vars-possible_tp)))
}
roc_df.one<-function(roc_analysis,thr){ # get the coordinates and cut offs for all the points in the ROC object
  roc_analysis<-roc_analysis[unlist(lapply(roc_analysis,is.null))==F]
  all<-lapply(roc_analysis,coords,x=thr,input="thr")

@@

@@

m.roc.table<-miseq.roc.table(sum.df,0.001,possible_tp,possible_vars)
knitr::kable(m.roc.table)

@@

## 2) Hiseq out of the box

### ROC

@@

h.table.df<-hiseq.roc.table(sum.df,0.001,possible_tp,possible_vars)

@@

```{r, phred_mapq filter}
sig.df<-subset(sum.df,p.val<0.001)

ggplot(subset(sig.df,category %in% c(T,F)),aes(x=Phred,y=MapQ,color=category))+geom_point(alpha=0.5)+facet_wrap(~gc)

ggplot(subset(sig.df,gc==5 & category %in% c(T,F)),aes(x=Phred,y=MapQ,color=category))+geom_point()

@@

h.cut.table<-hiseq.roc.table(cut.df,0.001,possible_tp,possible_vars)

knitr::kable(subset(h.cut.table,gc==4))
```

@@

### Read Position
```{r,read_pos}
ggplot(sig.df,aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
ggplot(subset(sig.df,gc==5),aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)

ggplot(cut.df,aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
ggplot(subset(cut.df,gc==5),aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)

sig.cut.df<-subset(cut.df,p.val<0.001)

ggplot(sig.cut.df,aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
ggplot(subset(sig.cut.df,gc==5),aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)

@@

h.cut.read.table<-hiseq.roc.table(cut.read.df,0.001,possible_tp,possible_vars)
knitr::kable(subset(h.cut.read.table,gc==5))
```
### Lofreq out of the box
```{r,lofreq}

@@

```
### Frequency accuracy
```{r, frequency_accuracy}
# Here we'll just be using

@@

## Figure 5 Lower inputs

```{r,roc lower_inputs}
ggplot(subset(h.cut.read.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))

ggplot(subset(h.cut.read.df,gc==3),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))
```

@@

```{r,duplicates}
sum.df<-read.csv("../data/process/2015-11-14/Variants/all.sum.csv",stringsAsFactors = F,comment.char = "#")
pat<-"([0-9]+)_([0-9]+)"
mutate(sum.df, exp.freq = sub(pat,"\\1",Id),
       dup = as.numeric(sub(pat,"\\2",Id)),
)-> sum.df
mutate(sum.df,exp.freq=ifelse(test = grepl("0",exp.freq),yes=as.numeric(exp.freq)/1000,no = as.numeric(exp.freq)/100))->sum.df

sum.df<-subset(sum.df,Id!="PC") # remove the plasmid control that was run

@@

dups<-mutate(dups,p.val=pmax(p.val1,p.val2),Id=gsub(pat,'\\1',Id),gc=4)

@@

dups.table<-hiseq.roc.table(dups,0.001,possible_tp,possible_vars)
knitr::kable(subset(dups.table))

@@

>>>>

---
title: "Figures"
author: "JT McCrone"
date: "December 30, 2015"
output: html_document
---

Here are the figures (or figure panels) we are thinking of using in the Benchmarking paper.  I will include my comments and decriptions on each and perhaps over time this will develop into the results section
```{r,settings_and_lib,echo=F,message=F,}
require(ggplot2)
require("plyr")
require("Biostrings")
require(reshape2)
require("pROC")
require(gridExtra)
theme_set(new = theme_classic(base_size=24))
require(knitr)
opts_chunk$set(fig.align = "center", warning = FALSE, echo = FALSE, fig.width = 10, 
    fig.height = 7,message=F)
```

## Figure 1

The accuracy of variants callers is most appropiately described by the sensitivity and specificity of a the caller at a given qualtiy threshold.  The sensitivity of a caller is the proportion of true, known variants identified by the algorith. The specificty is the the propotion of false variants corrected ignored. There is a trade off between sensitivity and specificity.  One may be able to correctly identify all known variants from a data; however, if in doing the the algorithm identifies a large number of false positives then a sensitivity of 1 is of little use in practive. Conversly it is possible to increase the stringency of a caller to the point that it does not identify any erroneous variants however, if in doing so all the true variants are also removed, perfect specificity is of no use.  A reciever operator curve (ROC) is a common way of visualizing the trade off between sensitivity and specificity. The DeepSNV algorithm returns a p value for each called variant. Figure 1 A depicts the relationship between various p value thresholds, and the resulting ROC using a hypothetical data set.

### ROC skematic
```{r, roc_skematic}
set.seed(42) # make the figures reproducible
example<-data.frame(Condition=c(rep(x = c(TRUE,FALSE),each=20)),test=c(runif(20,-10,-4),runif(20,-6,0))) # Make a dataframe of T and F positives with random p values that favors a lower distribution for the TP
ex.roc<-roc(example$Condition~example$test) # make an ROC based on this data

@@

#Helper function that gets the threshold and sensitivity from and roc at a given speceificity
helper.roc<-function(spec,roc.df){
  above.sense.cut<-subset(roc.df,specificity>=spec,select=c(threshold,specificity,sensitivity)) # get the points where specificity is above the cutoff
  sense.min<-min(above.sense.cut$specificity) # get the minimum of these to get as close to desired sensitivity as possible
  sense.cut<-subset(above.sense.cut,specificity==sense.min) # get all rows where sensitivity is at this point
  spec.max<-max(sense.cut$sensitivity) # but there can be different FPR for each TPR so here we maximize specificity
  output<-subset(sense.cut,sensitivity==spec.max)
}

@@

The accuracy of DeepSNV algorithm has been extensively benchmarked using a region of the HIV genome (?kb); however, we wanted to determine its accuracy in calling variants accross the entire influenza A genome ( 8 segments totally roughly 13kb).  To this end we diluted one lab strain WSN33 into another PR8 at decreasing frequencies. 491 single nucleotide variants were confirmed between these strains using sanger sequencing of plaque isolated samples.

### DeepSNV ROC 
```{r,roc_helper_runctions}
id<-function(x){
  x<-strsplit(x,".",fixed=T)[[1]][1]
}

@@

adjust.coords<-function(roc.df,sum.df,possible_tp,possible_vars){ # adjust the sensitivity and specificity called by pROC  
  samp<-roc.df$samp[1]
  samp.df<-subset(sum.df,Id==samp)
  sense.factor<-length(which(samp.df$category==T))/possible_tp
  TN.samp<-length(which(samp.df$category==F))
  mutate(roc.df,adj.sensitivity=sensitivity*sense.factor,FP=(TN.samp-TN.samp*specificity),TP=adj.sensitivity*possible_tp,adj.specificity=(possible_vars-possible_tp-FP)/((possible_vars-possible_tp)))
}
roc_df.one<-function(roc_analysis,thr){ # get the coordinates and cut offs for all the points in the ROC object
  roc_analysis<-roc_analysis[unlist(lapply(roc_analysis,is.null))==F]
  all<-lapply(roc_analysis,coords,x=thr,input="thr")

@@

The initial ROC of DeepSNV applied the viral dilutions above. Note for these analyses we only inlcude variants with p<0.05. I will extend this in the future. Doing so will extend the ROCs but will not change the results as we set a cut off of 0.01 at most in all the tables.

> Uniform ROC axis?

@@

m.roc.table<-miseq.roc.table(sum.df,0.01,possible_tp,possible_vars)
knitr::kable(m.roc.table)

@@

## Figure 2

### Hiseq expermintal setup

To test our accuracy under conditions that more closely resemble those found in patient derived samples, 20 single point mutants were mixed and diluted into wild type decreasing frequencies. This mixtures were then diltued into viral media to yeild a spectrum of known variants at a range of input levels.

### ROC
Out of the box here is how we do for the $10^5$ sample.

@@

h.table.df<-hiseq.roc.table(sum.df,0.01,possible_tp,possible_vars)

@@

It is common practice to account for base and mapping quality scores when calling variants. While DeepSNV is agnostic towards mapping qualty our initial analysis masked any base that had a phred score less than 30. Here we show that by focusing our analysis to varaints that have average phred >35 and average mapping quality >30 we can increase specificity without decrease sensitivity.

> The first plot here is just to show you the distributions. p<0.01. We usually only look at this after the read cut off. But here you can see the cut offs to quite a bit, but do leave a bit to do. the cut offs of 30 and 35 can be adjusted as well. There must be a nice way to find the cutoffs algorithmically.

```{r, phred_mapq filter}
sig.df<-subset(sum.df,p.val<0.01)

ggplot(subset(sig.df,category %in% c(T,F)),aes(x=Phred,y=MapQ,color=category))+geom_point(alpha=0.5)+geom_hline(yintercept=30)+geom_vline(xintercept=35)+facet_wrap(~gc)

#ggplot(subset(sig.df,gc==5 & category %in% c(T,F)),aes(x=Phred,y=MapQ,color=category))+geom_point()

@@

h.cut.table<-hiseq.roc.table(cut.df,0.01,possible_tp,possible_vars)

knitr::kable(subset(h.cut.table,gc==5))
```

This is still for p<0.01 and 10^5 genomes/ul viral transport media.

@@

### Read Position

> Again the first plot is just to show the distribution is the significant variants that pass the map and prhed cutoff for all samples run.

```{r,read_pos}
# ggplot(sig.df,aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
# ggplot(subset(sig.df,gc==5),aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
# 
# ggplot(cut.df,aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
# ggplot(subset(cut.df,gc==5),aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)

sig.cut.df<-subset(cut.df,p.val<0.01)

ggplot(sig.cut.df,aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)
# ggplot(subset(sig.cut.df,gc==5),aes(x=pos,y=Read_pos,color=category))+geom_point()+facet_wrap(~chr)

@@

h.cut.read.table<-hiseq.roc.table(cut.read.df,0.01,possible_tp,possible_vars)
knitr::kable(subset(h.cut.read.table,gc==5))
```

Same as above just with an added read cut off at 32 and 94 (middle 50%). p<0.01

### Lofreq out of the box

```{r,lofreq}

@@

### Frequency accuracy

For p<0.01 with phred, mapq and read position cut off.
```{r, frequency_accuracy}
# Here we'll just be using

@@

## Figure 5 Lower inputs
### 10^4
```{r,roc_lower_inputs4}
ggplot(subset(h.cut.read.df,gc==4),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))

knitr::kable(subset(h.cut.read.table,gc==4))
```
```{r,roc_lower_inputs3}
ggplot(subset(h.cut.read.df,gc==3),aes(x=1-adj.specificity,y=adj.sensitivity))+geom_step(aes(color=exp.freq),size=2)+scale_color_brewer(palette = "Set1")+xlab("1-Specificity")+ylab("Sensitivity")+guides(color=guide_legend(title="Frequency"))+scale_y_continuous(limits=c(0,1))

knitr::kable(subset(h.cut.read.table,gc==3))
```

@@

```{r,duplicates}
sum.df<-read.csv("../data/process/2015-11-14/Variants/all.sum.csv",stringsAsFactors = F,comment.char = "#")
#sum.df<-read.csv("~/Box Sync/Lab_work/NGS/Benchmarking/data/2015-11-14/all.sum.csv")
pat<-"([0-9]+)_([0-9]+)"
mutate(sum.df, exp.freq = sub(pat,"\\1",Id),
       dup = as.numeric(sub(pat,"\\2",Id)),
)-> sum.df

sum.df<-subset(sum.df,Id!="PC") # remove the plasmid control that was run
mutate(sum.df,Id=paste(4,exp.freq,sep="_"),exp.freq=ifelse(test = grepl("0",exp.freq),yes=as.numeric(exp.freq)/1000,no = as.numeric(exp.freq)/100))->sum.df

@@

dups<-mutate(dups,p.val=pmax(p.val1,p.val2))

dups<-subset(dups,p.val!=1)

@@

dups.table<-hiseq.roc.table(dups,0.01,possible_tp,possible_vars)
knitr::kable(subset(dups.table))

@@

These are variants that pass mapq,phred, and read cut and p<0.01 in both samples.10^4

This is so much better than before. Is there really that much variability between runs?

> tables are right but does the ROC make sense?

### Ordinations
```{r}
# No wt variants for this work. I'll start with the out of the box data p<0.01

sig.df<-subset(sig.df,category!="wt" & freq.var>0.01 & exp.freq>0.005) # no need to look at the samples were we expect things to be less than 1%

# at this point I'm just taking the counts on the forward direction and assuming that everyone has the same coverage this will be adjusted later to take a subsample

dcast(sig.df,Id~mutation,value.var ="freq.var",fill = 0)->counts

sample.names<-counts$Id
row.names(counts)<-sample.names
counts<-(subset(counts,select=-c(Id)))
vare.dist<-vegdist(counts);vare.dist

NMDS<-metaMDS(vare.dist,trymax = 100)
stressplot(NMDS)

plot(NMDS)
orditorp(NMDS,display="sites",cex=1.25,air=0.01)

cut.read.sig.df<-subset(cut.read.sig.df,category!="wt" & freq.var>0.01 & exp.freq>0.005) # no need to look at the samples were we expect things to be less than 1%

# at this point I'm just taking the counts on the forward direction and assuming that everyone has the same coverage this will be adjusted later to take a subsample

dcast(cut.read.sig.df,Id~mutation,value.var ="freq.var",fill = 0)->counts

sample.names<-counts$Id
row.names(counts)<-sample.names
counts<-(subset(counts,select=-c(Id)))
vare.dist<-vegdist(counts);vare.dist

NMDS<-metaMDS(vare.dist,trymax = 100)
stressplot(NMDS)

plot(NMDS)
orditorp(NMDS,display="sites",cex=1.25,air=0.01)
